---
title: "Log crossing activity in a tropical rainforest"
author: "Juan S. Vargas Soto"
output: html_document
bibliography: references.bib
editor_options: 
  chunk_output_type: inline
---

```{r global options, include=FALSE}
knitr::opts_chunk$set(message = FALSE, echo = FALSE)
```

```{r load libraries, echo=FALSE}
library(tidyverse)
library(lubridate)
library(lme4)
library(performance)
```

## Summary

We analyzed how different species of mammals use logs fallen across rivers in a tropical wet forest. We found that most species observed use logs to cross rivers, but there are significant differences among species, both in the frequency of detection at these logs, and the probability that they will cross; some are much more likely to use them. While some factors like the length of the log, its width, and how high it is above the water may increase this probability, these effects are small compared to baseline differences among species. Also, seasonality does not appear to play an important role in the probability of use of the log, although many of our cameras were only working during the rainy season so this factor is not clear yet. 
The species that use logs to cross most frequently are species that regularly use tree trunks and branches like coatis, opossums, and ocelots, so crossing these bridges should not represent an important risk. Some species that did not cross the logs, like pacas and tapirs, were nonetheless observed walking in the water, so it is possible that they are also able to cross rivers regularly despite not using logs. This is not the case for agoutis, which had a low probability of crossing, and were also never observed in the river. For agoutis, rivers may therefore represent significant barriers that limit their home range. Considering the diversity and frequency of use of these logs, they may define important linear features in the forest. As such, they could represent also important sites for communication, as seen by the frequent scent-marking, for male coatis in particular.

## Introduction

The goal of this project was to study how trees that have fallen over rivers and streams in tropical forests---effectively creating a temporary bridge---are used by different species of terrestrial vertebrates, focusing particularly on medium to large mammals and ground birds. In tropical wet forests, rivers and streams may constitute significant barriers for movement, limiting the dispersal of animals. However, these forests also have frequent tree falls, which can happen by rivers and connect both sides. Here, we aim to determine if these fallen trees effectively serve as bridges, and which species of animals are able to use them.

## Methods

We searched for logs across rivers throughout the Osa Wildlife Refuge in southern Costa Rica. We placed motion-activated cameras overlooking the logs, and recorded all activity. For every animal detected, we identified the species, whether or not it crossed the log, and annotated any particular behavior, such as scent-marking. Records were organized and annotated using the Timelapse software (@Greenberg2019). We set up 26 cameras overall at 18 different log (8 logs had cameras on both ends). If we ignore a camera that malfunctioned (CLOG6-B), cameras were working for over 6 months on average. Three cameras were working for more than a year (CLOG6-A, PLOG5-A,B), one for nearly 2 years (CLOG3-A).

```{r preproc, echo=FALSE}
# Import data and preprocess
dat_raw <- read_csv("../Data/timelapse_database.csv")
# Filter to keep only unique records. Make sure to arrange it so that if there
# is a crossing do not delete it.
dat <- dat_raw %>% 
  mutate(station = if_else(str_length(station)==5, str_replace(station, "G", "G0"), station)) %>% # Fix station names
  filter(!species %in% c("nothing", "setup", "corrupted","uid", 
                                      "lizard_uid","iguana_green", "bat", "bird", "horse", 
                                      "people", "dog")) %>% # remove non-animal records
  group_by(station, species) %>% arrange(station, species, DateTime) %>% 
  mutate(tdif = difftime(DateTime,lag(DateTime),units = "mins")) %>% # time difference between detections
  mutate(newevent = is.na(tdif) | tdif>duration(5, "minutes")) %>% # is the detection a new event?
  ungroup(species) %>% 
  mutate(eventnum = cumsum(newevent)) %>% # label independent events
  group_by(station,eventnum) %>% mutate(eventcross = any(crossing)) %>% ungroup() %>% 
  filter(newevent) %>% select(station, species, DateTime, eventcross, behavior)

# Covariates data import and merge with detections
log_data <- readxl::read_excel("../Data/covariates_data_base.xlsx")
names(log_data)[1] <- "station"
dat_wcovs <- dat %>% left_join(log_data)
# scale covariates for model fitting, join with detection data
log_data_scaled <- mutate(log_data, across(Log_lenght:Vel_conver,datawizard::standardise))
dat_wcovs_scaled <- left_join(dat,log_data_scaled)
```
## Results

### Effort
The total effort was 2374 trap-days, the shortest effort was 12 days at a station, and the longest 342 (counting both cameras together).

```{r}
effortdb <- read.csv("../Data/effort.csv")
effortdb %>%  group_by(station, camera) %>% summarise(cumeffort = sum(effort)) 
```

```{r}
dat_raw %>% filter(!DeleteFlag) %>% 
  ggplot(aes(DateTime, station))+
  geom_area(data = data.frame(station = "PLOG8", DateTime = strptime(c("2020-05-01","2020-11-30"), format = "%Y-%m-%d")), fill = "lightblue")+
  geom_area(data = data.frame(station = "PLOG8", DateTime = strptime(c("2021-05-01","2021-11-30"), format = "%Y-%m-%d")), fill = "lightblue")+
  geom_point(aes(color = camera),position = position_dodge(width = 0.5))+
  geom_point(aes(color = camera), shape = 2, data = \(x) filter(x,setup), position = position_dodge(width = 0.5))
```
We see that the effort was distributed in two periods. The first was between June and October 2020, all within the rainy season. In 2021, only five stations were deployed again, starting in March 2021. These cameras would have had two months of dry season, and the beginning of the rainy season in May/June. This is important to account for the total effort. I've created a file with the dates for the different periods. 

### Species observed

We identified 22 species of wild mammals. We also recorded bats and rodents, but these groups are not consistently detected and cannot be properly identified. There were several species of birds, but we keep only one species that is mostly terrestrial for further analyses, the great Curassow *Crax rubra*. Finally, we observed people (82 detections) and domestic animals (5 detections of dogs and 4 of horses) but we do not include these in further analyses. The most common species were tamanduas (*Tamandua mexicana*), followed by coatis (*Nasua narica*), common opossums (*Didelphis marsupialis*), striped skunks (*Conepatus semistriatus*), ocelots (*Leopardus pardalis*), tayras (*Eira barbara*), curassows, pumas (*Puma concolor*), agoutis (*Dasyprocta punctata*), squirrels (*Sciurus granatensis*), pacas (*Cuniculus paca*), and four-eyed opossums (*Philander opossum*). All other species were detected fewer than 10 times. Species did not use the logs equally. For opossums, ocelots, tamanduas, coatis, tayras, pumas, and skunks, crossings constituted over 80% of all sightings. In contrast, species like agoutis and curassows crossed in fewer than 50% of sightings. Species like pacas and peccaries were never seen crossing logs. Raccoons (n = 10), and tapirs (n=1) were detected few times, but were occasionally seen walking in the river. These observations were only possible for a few stations that had a vantage point overlooking both the log and the river.

```{r det summary, echo=FALSE}
dat %>% count(species, eventcross) %>% 
  pivot_wider(names_from = eventcross, values_from = n, values_fill = 0) %>% 
  mutate(detections = `FALSE`+`TRUE`,prop = 100*`TRUE`/(`FALSE`+`TRUE`)) %>%
  select(species, detections, prop) %>% 
  arrange(desc(detections)) %>% 
  knitr::kable(digits = 1, col.names = c("Species","Detections (n)", "Cross freq. (%)"), caption = "Summary of detections of different species of mammals and birds around logs over rivers")
```

There seem to be some important differences in the probability of crossing across species, some cross nearly every occasion while others rarely do. These observations, however, must be analyzed only for cameras where the same bank is visible.

```{r det summary 2, echo=FALSE, fig.cap="Activity and crossing probability for wild vertebrates near logs above rivers in the Osa Wildlife Refuge, Costa Rica"}
dat %>% group_by(species) %>% 
  filter(n()>10) %>% 
  group_by(station, species) %>% 
  mutate(prop = 100*sum(eventcross)/n()) %>% 
  ggplot()+geom_count(aes(fct_reorder(species, prop), station, color = prop))+
  # summarise(prop = sum(eventcross)/n()) %>% 
  # ggplot()+geom_raster(aes(y = station, x = species, fill = prop))+
  theme_classic(base_size = 12)+
  theme(axis.text.x = element_text(hjust = 0, angle = -45))+
  scale_color_continuous(type="viridis")+
  labs(x = "Species", y = "Station", color = "Cross proportion (%)")
```

We see that the proportion of crossings is mostly consistent across logs for the same species, as seen from the color range observed vertically, with maybe the exception of tayras. This proportion varies mainly across species. The diversity of species at each log is also different. Logs like PLOG05, PLOG06, and PLOG02 detected 15, 14, and 13 species, respectively. PLOG01 and PLOG08 in contrast each recorded only three different species.

### Factors influencing use of logs

We were interested in determining factors that could influence the activity at the log and the probability of crossing. We measured different variables about the log (width, length, height above the water), the river or stream characteristics at that spot (width, depth, flow), and the amount of vegetation (canopy cover). Given the number of covariates, we attempted to reduce it to fewer dimensions by testing for significant correlations among them, and by determining the principal components.

```{r cov correlation, echo=FALSE, fig.cap="Correlation among covariates for logs across rivers in the Osa Wildlife Refuge"}
covs_cor <- log_data %>% select(-c(1,9,16)) %>% cor()
covs_cor_p <- log_data %>% select(-c(1,9,16)) %>% corrplot::cor.mtest() 
corrplot::corrplot(covs_cor,order = "hclust", p.mat = covs_cor_p$p, tl.col = "black", insig = "blank", diag = F, type = "lower", tl.srt = 45)
```

The covariates were for the most part not strongly correlated to each other. There were some positive correlations between stream width, canopy cover and open percentage. There was also a negative correlation between stream width, canopy cover, and percentage open, and bark. Additionally, stream velocity was positively correlated with log diameter and height above water, and negatively correlated with the velocity of the stream. There are nonetheless many covariates, and it could make sense to reduce the dimensions using a principal components analysis (PCA).

```{r covar pca, echo=FALSE, fig.cap = "Principal component analysis of covariates for logs over rivers in the Osa Wildlife Refuge"}
pca <- log_data %>% select(2:8,10) %>% prcomp(scale. = T)
par(mfrow = c(1,2))
biplot(pca)
screeplot(pca)
```

```{r pca summary, echo=FALSE, include=FALSE}
summary(pca)
```

It may be hard to summarize the variables in fewer than 3 components. Because of this it makes more sense to keep the original variables, and select covariates to include based on relevance and model fit. Here are the variables that I believe should be relevant:

-   Log diameter

-   Stream width

-   Height above water

-   Seasonality for frequency and probability of crossing

Below is the average montly precipitation in Pto. Jimenez, which we can use as a continuous variable instead of the season.

```{r}
rainfall = data.frame(month = 1:12,rain = c(77,53,65,141,370,357,373,427,411,627,593,226))
rainfall <- mutate(rainfall,rain_sc = scale(rain))
```

### Number of species crossing

There doesn't seem to be any factor that determines the number of species crossing at each station. The only one where it could seem there is an important effect is the log diameter, where wider logs increase the number of species that cross it.

```{r}
d1 <- dat %>% filter(eventcross) %>% count(station,species) %>% count(station) %>% 
  left_join(log_data_scaled)
m1 <-glm(n~Log_to_water+Diameter+Stream_width+River_depth, data=d1, family = poisson(link = 'log')) 
par(mfrow=c(2,2))
plot(m1)
```

```{r}
summary(m1)
```



### Crossing frequency

There were 1066 independent events, so 1096/2374 trap days = 0.45 /day, but this varied across stations, between 0.05 and 1.1 events/day.

```{r}
dat %>% group_by(station) %>% left_join(summarise(effortdb, effort = sum(effort),.by = station)) %>% summarize(n = n(), freq = n()/mean(effort),prop = sum(eventcross)/n())
```

We analyze the number of animals seen crossing. For this, we divide the detection into weekly detection frequencies. The weeks that the camera was working are in a csv file. 

Below I create a database with the number of weekly detection for every station and species

```{r}
weekeffort <- dat %>% group_by(species) %>% filter(n()>10, eventcross) %>% distinct(station,species) %>% full_join(read.csv("../Data/station_weeks.csv"))

# Create database for model
dat_glmm_week_freq <- dat %>% group_by(species) %>% filter(n()>10, eventcross) %>% ungroup() %>% 
  mutate(wk = week(DateTime), yr = year(DateTime)) %>% 
  count(station,species,yr,wk) %>% 
  right_join(weekeffort) %>% arrange(station,yr,wk) %>% 
  mutate(n = replace_na(n,0)) %>% 
  mutate(month = ceiling(wk/4.3), season = if_else(wk>16&wk<48, "rainy","dry")) %>% 
  left_join(rainfall) %>% left_join(log_data_scaled) 
```
With the database, we can run the models. First I will find the most appropriate random effect structure. The base frequency (intercept) should be related to the species, some are just more abundant than others. Previous studies have also found that seasonality impacts different species differently, so this can also be included as a random effect. For this first step I will include all covariates of interest, i.e. the stream width and depth, log height, log width, log length, and season. The response variable is the number of crossings by species per week. 
```{r}
freqmodels <- list()
# run model. Mixed effects model of number of detections per week wrt covariates. 
freqmodels$full_RndIntSsn <- glmer(n ~ Log_lenght+Diameter+Log_to_water+River_depth+season+(1+season|species), data = dat_glmm_week_freq, family = poisson)
freqmodels$full_RndInt <- update(freqmodels$full_RndIntSsn, ~ Log_lenght+Diameter+Log_to_water+River_depth+season+(1|species))
freqmodels$full_NoRnd <- glm(n ~ Log_lenght+Diameter+Log_to_water+River_depth+season, data = dat_glmm_week_freq, family = poisson)

AIC(freqmodels[[1]],freqmodels[[2]], freqmodels[[3]])
```
The best structure seems to be the model with both season and species as random effects. In this model, all covariates have a clear effect on the number of detections. There would be fewer crossings for deeper rivers, and more crossings for longer, wider logs higher above the water. 

```{r}
summary(freqmodels$full_RndIntSsn)
```



```{r}
ranef(freqmodels$full_RndIntSsn) |> dotplot.ranef.mer()
```

The random species intercept estimates suggest that four-eyed opossums have lower baseline activity than the rest of the species, and that common opossums and coatis have higher than average. The season seems to only affect curassows and coatis. Their use of logs would decrease in rainy season.

```{r}
check_zeroinflation(freqmodels$full_RndIntSsn)
check_overdispersion(freqmodels$full_RndIntSsn)
```

The model seems to be zero-inflated and overdispersed. It is recommended to use a zero-inflated Poisson or zero-inflated negative binomial distribution instead.

#### Zero-inflated model

First I specify a zero-inflated Poisson mixed model. This model requires specifying a model for the zero-inflation process, i.e. the process linked to additional 0s in the observed data. For simplicity I set this model as a intercept-only model, so the additional 0s will be linked to a constant. This assumes that zero-inflation is uniform across stations and species.

```{r}
library(glmmTMB)
zifreqmodel <- glmmTMB(formula(freqmodels$full_RndIntSsn), data = dat_glmm_week_freq, family = poisson, ziformula = ~1)
zifreqmodel2 <- update(zifreqmodel, ~.-(1+season|species)+(1|species))
AIC(zifreqmodel, zifreqmodel2, freqmodels$full_RndIntSsn)
```
The best model in this case is the one with both random effects (intercept and season). Both zero-inflated models have a significantly lower AIC than the Poisson model

```{r}
check_zeroinflation(zifreqmodel)
check_overdispersion(zifreqmodel)
```
This model still has overdispersion and zero-inflation. Here I include covariates in the zero-inflation model. I attempt to include the station or the species as a random effect.

```{r}
zifreqmodel3 <- update(zifreqmodel, ziformula = ~ (1|station))
zifreqmodel4 <- update(zifreqmodel, ziformula = ~ (1|species))
AIC(zifreqmodel,zifreqmodel2, zifreqmodel3, zifreqmodel4)
```
Out of these, the best model according to AIC is the one that includes station as a random effect in the binomial false zero model. 

```{r}
check_zeroinflation(zifreqmodel3)
check_overdispersion(zifreqmodel3)
```
This model continues to be overdispersed and zero-inflated. Next let's build the same zero-inflated models with a negative binomial distribution instead of Poisson.

```{r}
freqmodel_nb <- glmmTMB(n ~ Log_lenght+Diameter+Log_to_water+River_depth+season+Stream_width+(1+season|species), data = dat_glmm_week_freq, family = nbinom1)
freqmodel_nb2 <- update(freqmodel_nb, ~Log_lenght+Diameter+Log_to_water+River_depth+season+Stream_width+(1|species))
zinbfreqmod <- glmmTMB(formula = formula(freqmodel_nb), data = dat_glmm_week_freq, family = nbinom1, ziformula = ~ 1)
zinbfreqmod2 <- glmmTMB(formula = formula(freqmodel_nb2), data = dat_glmm_week_freq, family = nbinom1, ziformula = ~ 1)
AIC(freqmodel_nb2, freqmodel_nb, zinbfreqmod2, zinbfreqmod)
```
All these models have lower AIC than the ones with a Poisson distribution. Out of them, the best model is the one with negative binomial distribution and season as a fixed effect. 

```{r}
check_zeroinflation(freqmodel_nb2)
check_overdispersion(freqmodel_nb2)
```
This model is still overdispersed but not zero-inflated.  
```{r}
bestfreqmod <- freqmodel_nb2
summary(bestfreqmod)
```
SOOOO, THE FINAL DECISION (MAYBE). I fit a negative binomial mixed effects model with the number of detections as a response variable. I included a random intercept species.  I included the log length, log width, height above water, stream width, stream depth, and season as fixed effects. 


```{r}
predict(bestfreqmod, newdata = data.frame(Log_lenght=0, Diameter=0, Log_to_water=0, River_depth=0, Stream_width=0, season = 'dry'), type='response', se.fit = T, re.form = NA)
```
```{r}
icc(bestfreqmod)
```
The ICC of 0.10 indicates some small correlation within species, so it may not be necessary to include the random effect.

#### Plot
```{r}
mumean <- predict(bestfreqmod, newdata = data.frame(Log_lenght=0, Stream_width=0, season="dry", Log_to_water=0, River_depth=0, Diameter=0), type = "response", re.form=NA)
as.data.frame(ranef(bestfreqmod)) %>% 
  mutate(int = fixef(bestfreqmod)[[1]][1]) %>% 
  mutate(est = exp(condval+int)/(1+exp(condval+int)),
         hi = exp(int+condval+2*condsd)/(1+exp(int+condval+2*condsd)),
         lo = exp(int+condval-2*condsd)/(1+exp(int+condval-2*condsd))) %>% 
  # mutate(sp = fct_relabel(grp, )) %>% 
    left_join(read.csv("../Data/spnames.csv"), by = join_by(grp==species)) %>% 
  ggplot(aes(fct_reorder(plotname, est),est)) +
  geom_hline(yintercept = mumean, linetype=2) +
        geom_point() + 
        geom_errorbar(aes(ymin=lo,
                           ymax=hi), width=0)+
  theme_classic(base_size = 14)+
  labs(x = "Species", y = "Use frequency (crossings/week)")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = -35, hjust = 0),
        plot.margin = unit(c(1,5,1,1), "mm"), 
        aspect.ratio = 0.9,
        panel.border = element_rect(fill=NA))



basefreqpred <- predict(bestfreqmod, newdata = data.frame(Log_lenght=0, Stream_width=0, season="rainy", Log_to_water=0, River_depth=0, Diameter=0, species=unique(dat_glmm_week_freq$species)), type = "response", se.fit = T)
pfreqs <- data.frame(freq = basefreqpred$fit, se = basefreqpred$se.fit, species = unique(dat_glmm_week_freq$species)) %>% 
  left_join(read.csv("../Data/spnames.csv")) %>% 
  ggplot()+
  geom_hline(yintercept = mumean, lty=2)+
  geom_hline(yintercept = 0)+
  geom_pointrange(aes(y=freq, ymin = freq-2*se, ymax=freq+2*se, x=fct_reorder(plotname, freq)))+
  labs(x = "Species", y = "Use frequency (crossings/week)")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = -45, hjust = 0),
        plot.margin = unit(c(1,5,1,1), "mm"), 
        panel.border = element_rect(fill=NA))
pfreqs
```

```{r}
og_vars_meansd <- cbind(apply(select(log_data, where(is.numeric)), 2, mean),
                        apply(select(log_data, where(is.numeric)), 2, sd))
pspnewdata <- data.frame(Log_to_water = seq(min(bestfreqmod$frame$Log_to_water),
                                            max(bestfreqmod$frame$Log_to_water),
                                            length.out = 100), 
           Diameter = 0, 
           Log_lenght = 0, 
           Stream_width = 0, 
           River_depth = 0,
           season = "rainy") %>% 
  mutate(Log_to_water_og = og_vars_meansd["Log_to_water",1]+og_vars_meansd["Log_to_water",2]*Log_to_water)
ypred <- predict(bestfreqmod, type = "response", newdata = pspnewdata, se.fit = T, re.form = NA) 
pht <- cbind(pspnewdata, mu=ypred$fit, se=ypred$se.fit) %>% ggplot(aes(Log_to_water_og, mu))+
  geom_ribbon(aes(ymin = mu-se, ymax = mu+se), fill = "gray80")+
  geom_line()+
  theme_classic(base_size = 12)+
  labs(x = "Height above water (m)", y="Crossing frequency (/week)")
pspnewdata <- data.frame(Log_lenght = seq(min(bestfreqmod$frame$Log_lenght),
                                            max(bestfreqmod$frame$Log_lenght),
                                            length.out = 100), 
           Diameter = 0, 
           Log_to_water = 0, 
           Stream_width = 0, 
           River_depth = 0,
           season = "rainy") %>% 
  mutate(Log_lenght_og = og_vars_meansd["Log_lenght",1]+og_vars_meansd["Log_lenght",2]*Log_lenght)
ypred <- predict(bestfreqmod, type = "response", newdata = pspnewdata, se.fit = T, re.form = NA) 
plen <- cbind(pspnewdata, mu=ypred$fit, se=ypred$se.fit) %>% ggplot(aes(Log_lenght_og, mu))+
  geom_ribbon(aes(ymin = mu-se, ymax = mu+se), fill = "gray80")+
  geom_line()+
  theme_classic(base_size = 12)+
  labs(x = "Log length (m)", y="Crossing frequency (/week)")
pspnewdata <- data.frame(Diameter = seq(min(mod$frame$Diameter),
                                            max(mod$frame$Diameter),
                                            length.out = 100), 
           Log_to_water=0,
           Log_lenght = 0, 
           Stream_width = 0, 
           River_depth = 0,
           season = "rainy") %>% 
  mutate(Diameter_og = og_vars_meansd["Diameter",1]+og_vars_meansd["Diameter",2]*Diameter)
ypred <- predict(bestfreqmod, type = "response", newdata = pspnewdata, se.fit = T, re.form = NA) 
pwid <- cbind(pspnewdata, mu=ypred$fit, se=ypred$se.fit) %>% ggplot(aes(Diameter_og, mu))+
  geom_ribbon(aes(ymin = mu-se, ymax = mu+se), fill = "gray80")+
  geom_line()+
  theme_classic(base_size = 12)+
  labs(x = "Log width (cm)", y="Crossing frequency (/week)")
pspnewdata <- data.frame(River_depth = seq(min(mod$frame$River_depth),
                                            max(mod$frame$River_depth),
                                            length.out = 100), 
           Log_to_water = 0,
           Diameter = 0, 
           Log_lenght = 0, 
           Stream_width = 0, 
           season = "rainy") %>% 
  mutate(River_depth_og = og_vars_meansd["River_depth",1]+og_vars_meansd["River_depth",2]*River_depth)
ypred <- predict(bestfreqmod, type = "response", newdata = pspnewdata, se.fit = T, re.form = NA) 
pdep <- cbind(pspnewdata, mu=ypred$fit, se=ypred$se.fit) %>% ggplot(aes(River_depth_og, mu))+
  geom_ribbon(aes(ymin = mu-se, ymax = mu+se), fill = "gray80")+
  geom_line()+
  theme_classic(base_size = 12)+
  labs(x = "River depth (cm)", y="Crossing frequency (/week)")
pspnewdata <- data.frame(Stream_width = seq(min(mod$frame$Stream_width),
                                            max(mod$frame$Stream_width),
                                            length.out = 100), 
           Log_to_water = 0,
           Diameter = 0, 
           Log_lenght = 0, 
           River_depth = 0, 
           season = "rainy") %>% 
  mutate(Stream_width_og = og_vars_meansd["Stream_width",1]+og_vars_meansd["Stream_width",2]*Stream_width)
ypred <- predict(bestfreqmod, type = "response", newdata = pspnewdata, se.fit = T, re.form = NA) 
prwid <- cbind(pspnewdata, mu=ypred$fit, se=ypred$se.fit) %>% ggplot(aes(Stream_width_og, mu))+
  geom_ribbon(aes(ymin = mu-se, ymax = mu+se), fill = "gray80")+
  geom_line()+
  theme_classic(base_size = 12)+
  labs(x = "River width (m)", y="Crossing frequency (/week)")
pspnewdata <- data.frame(
  River_depth = 0, 
           Log_to_water = 0,
           Diameter = 0, 
           Log_lenght = 0, 
           Stream_width = 0, 
           season = c("rainy","dry")
)
ypred <- predict(bestfreqmod, type = "response", newdata = pspnewdata, se.fit = T, re.form = NA) 
pses <- cbind(pspnewdata, mu=ypred$fit, se=ypred$se.fit) %>% 
  ggplot(aes(season, mu))+
  geom_pointrange(aes(ymin = mu-se, ymax = mu+se))+
  theme_classic(base_size = 12)+
  scale_x_discrete(labels = c("Dry","Rainy"))+
  labs(x = "Season", y="Crossing frequency (/week)")
plot_grid(pfreqs,plot_grid(plen,pwid,pht,pdep,prwid,pses, align = 'hv', labels = 'auto'))
```

### Probability of crossing

To analyze how the probability of crossing is influenced by different covariates, we used a binomial generalized linear model (GLM), where the response variable was whether the animal crossed or not for every observation, and with some selected covariates as predictors. In addition to the variables seen previously, I will also include the season, as rain will have significant effects on water level and flow. For one, the rain affects the water level, some streams will become easier to pass, rendering the logs unnecessary. Secondly, animals change their activity ranges in rainy and dry seasons, so there may be differences in overall activity near the streams. There are some clear differences between the species in their behavior and whether they tend to cross logs or not. We include a random effect of species to account for these differences. For this model, we include only species that had 10 or more records over the study period. In this analysis we only use cameras were you can see enough of the proximate bank to detect animals that do not cross. Most cameras were directed towards the middle of the log, so they would only be triggered once an animal was on the log.

```{r mm fit, echo=FALSE, fig.height=7, fig.cap="Diagnostic plots for mixed effects binomial model of crossing probability"}
library(lme4)
# Add the season to the database
# filter to keep only species with more than 10 records
dat_glmm <- dat_wcovs_scaled %>% filter(station %in% c("CLOG03", "CLOG07", "CLOG08", "CLOG09", "PLOG02", "PLOG04")) %>% # only keep stations overlooking the near bank
  mutate(month = month(DateTime), season = if_else(lubridate::month(DateTime) %in% c(1:4,12),"dry","rainy")) %>% left_join(rainfall) %>% 
  group_by(species) %>% filter(n()>10) %>% ungroup()
# Build model with season as a covariate, a random effect of species, and multiple covariates
mm1 <- glmer(eventcross~Log_to_water+Diameter+Log_lenght+Stream_width+River_depth+season+(1|species), data=dat_glmm, family = "binomial")
# model without season or rain
mm2 <- glmer(eventcross~Log_to_water+Diameter+Log_lenght+Stream_width+River_depth+(1|species), data=dat_glmm, family = "binomial")
#compare the two
AIC(mm1,mm2)
```

The first model, with season, performs better than the one without.

```{r}
performance::check_model(mm1)
```

The model performs quite poorly, there is significant VIF, probably due to the low number of logs. Let's reduce the number of variables.

```{r}
mm3 <- glmmTMB(eventcross~Diameter+season+(1|species), data=dat_glmm, family = "binomial")
check_model(mm3)
```

```{r}
summary(mm3)
```
The best performing model is the one with only Diameter and season, and only season has a significant effect. The likelihood of crossing is higher during the rainy season.

#### Baseline crossing probability

```{r}
basecrossprob <- predict(mm3, newdata=data.frame(season=c('rainy',"dry"), Diameter=0), se.fit=T,type='response', re.form = NA)
```

The baseline probability for all species is high, around 0.75, but it also depends on the species. 

```{r}
ranef(mm3) |> dotplot.ranef.mer()
```

#### Effect of species

```{r ranef, echo=FALSE, fig.cap="Random effect of species on the intercept in the mixed effects model of crossing probability"}
spp <- unique(mm3$frame$species)
pspnewdata <- expand.grid(
  Diameter = 0, 
  season = c("rainy","dry"),
  species = spp
)
ypred <- predict(mm3, type = "response", newdata = pspnewdata, se.fit = T) 
cbind(pspnewdata, mu=ypred$fit, se=ypred$se.fit) %>% 
    left_join(read.csv("../Data/spnames.csv")) %>% 
  ggplot(aes(fct_reorder(plotname, mu), mu))+
  geom_pointrange(aes(ymin = mu-se, ymax = mu+se, color=season), position = position_dodge(0.3))+
  scale_color_manual(values = c("dodgerblue","gold2"), labels = c("Rainy","Dry"))+
  theme_classic(base_size = 16)+
  geom_hline(yintercept = basecrossprob$fit, linetype=2, color = c("dodgerblue","gold2"))+
  theme(axis.text.x = element_text(angle = -45, hjust=0))+
  labs(x = "Season", y="Crossing probability", color="Season")
```

The random effect shows that for about half of the species, the intercept is not clearly different from the mean intercept. For ocelots, common opossums, coatis, and tamanduas, their base probability of crossing will be higher than the mean effect. Meanwhile, curassows, agoutis, and pacas, have lower probabilities than the mean. The mean base probability of crossing is 83%, taking the median values for all the covariates. Transformed to base probabilities, we see that there are large differences, for coatis, opossums, and ocelots, the base probability of crossing is greater than 70%, while for agoutis it is less than 20%, and practically zero for pacas.

The confidence intervals for the model indicate that the inter-species variance in the intercept is significantly different from 0, and should be considered explicitly. 

#### Effect of covariates

```{r margins}
marfx <- margins::margins(mm1)
plot(marfx)
```

```{r}
summary(marfx)
```

The probability of crossing increases by 2.3% for an increase in 1 m in the height above the stream. The probability increases by 0.2% for an additional cm in log diameter. An 1 m increase in log length results in a 1.6% increase in probability. An increase of 1% in canopy cover results in an decrease of 3.2% in the probability. Finally, the probability decreases by 2.2% for every additional centimeter of stream depth. The plots below show how these effects would change the average probability of crossing (regardless of species). For all the covariates there is about a 15% increase or decrease in probability across the range of values observed in the study. Nevertheless, the greatest effects are those given by the species.

```{r effect plots, echo=FALSE, fig.cap="Effects of environmental covariates on the probability of crossing logs across rivers"}
newdata <- data.frame(Log_to_water = seq(min(mm2@frame$Log_to_water), max(mm2@frame$Log_to_water), length.out = 100), 
           Diameter = mean(mm2@frame$Diameter), 
           Log_lenght = mean(mm2@frame$Log_lenght), 
           Canopy_per = mean(mm2@frame$Canopy_per), 
           River_depth = mean(mm2@frame$River_depth))
ypred <- predict(mm2, type = "response", newdata = newdata, re.form = NA) 
p1 <- cbind(newdata, ypred) %>% ggplot(aes(Log_to_water, 100*ypred))+geom_line()+
  lims(y = c(0,100))+theme_classic(base_size = 12)+
  labs(x = "Height above water (m)", y="Crossing probability (%)")
newdata <- data.frame(Canopy_per = seq(min(mm2@frame$Canopy_per), max(mm2@frame$Canopy_per), length.out = 100), 
           Diameter = mean(mm2@frame$Diameter), 
           Log_lenght = mean(mm2@frame$Log_lenght), 
           Log_to_water = mean(mm2@frame$Log_to_water), 
           River_depth = mean(mm2@frame$River_depth))
ypred <- predict(mm2, type = "response", newdata = newdata, re.form = NA) 
p2 <- cbind(newdata, ypred) %>% ggplot(aes(Canopy_per, 100*ypred))+geom_line()+
  lims(y = c(0,100))+theme_classic(base_size = 12)+
  labs(x = "Canopy cover (%)", y="Crossing probability (%)")
newdata <- data.frame(Diameter = seq(min(mm2@frame$Diameter), max(mm2@frame$Diameter), length.out = 100), 
           Canopy_per = mean(mm2@frame$Canopy_per), 
           Log_lenght = mean(mm2@frame$Log_lenght), 
           Log_to_water = mean(mm2@frame$Log_to_water), 
           River_depth = mean(mm2@frame$River_depth))
ypred <- predict(mm2, type = "response", newdata = newdata, re.form = NA) 
p3 <- cbind(newdata, ypred) %>% ggplot(aes(Diameter, 100*ypred))+geom_line()+
  lims(y = c(0,100))+theme_classic(base_size = 12)+
  labs(x = "Log diameter (cm)", y="Crossing probability (%)")
newdata <- data.frame(River_depth = seq(min(mm2@frame$River_depth), max(mm2@frame$River_depth), length.out = 100), 
           Canopy_per = mean(mm2@frame$Canopy_per), 
           Log_lenght = mean(mm2@frame$Log_lenght), 
           Log_to_water = mean(mm2@frame$Log_to_water), 
           Diameter = mean(mm2@frame$Diameter))
ypred <- predict(mm2, type = "response", newdata = newdata, re.form = NA) 
p4 <- cbind(newdata, ypred) %>% ggplot(aes(River_depth, 100*ypred))+geom_line()+
  lims(y = c(0,100))+theme_classic(base_size = 12)+
  labs(x = "River depth (cm)", y="Crossing probability (%)")
newdata <- data.frame(Log_lenght = seq(min(mm2@frame$Log_lenght), max(mm2@frame$Log_lenght), length.out = 100), 
           Canopy_per = mean(mm2@frame$Canopy_per), 
           River_depth = mean(mm2@frame$River_depth), 
           Log_to_water = mean(mm2@frame$Log_to_water), 
           Diameter = mean(mm2@frame$Diameter))
ypred <- predict(mm2, type = "response", newdata = newdata, re.form = NA) 
p5 <- cbind(newdata, ypred) %>% ggplot(aes(Log_lenght, 100*ypred))+geom_line()+
  lims(y = c(0,100))+theme_classic(base_size = 12)+
  labs(x = "Log length (m)", y="Crossing probability (%)")
cowplot::plot_grid(p1,p3,p5, p2,p4)
```

```{r}
margins::margins_summary(mm2) %>% mutate(AME=AME+1)
```

Export combined figure

```{r combfig}
library(cowplot)
png("figures/covar_effects.png",width = 21,height = 9, units = 'cm',res = 300,pointsize = 12,bg = 'white')
plot_grid(psp,plot_grid(p1,p3,p5, p2,p4, align = 'hv',labels=c('b','c','d','e','f'), label_x = -0.05),rel_widths = c(0.4,0.6), labels = 'a',label_x = -0.05)
dev.off()
```

### Noteworthy behaviors

We see that many species use the logs, at multiple stations more than 10 different species were observed. This means that these features may be acting also as ways of communicating across and within species. In fact, we observed coatis (n = 23), common opossums (n = 1), striped skunks (n = 17), scent-marking on the logs. We also observed tamanduas defecating on logs on two occasions. It could be worth exploring whether these logs are important marking sites for territorial animals, and whether scent-marking prevents use by conspecifics. In addition to scent-marking sites, logs may act as a food source. We observed tamanduas and striped skunks foraging by removing the tree bark and searching for ants underneath, 10 times for tamanduas, but only once for skunks. We have observed this behavior in previous studies in the same area, so it is unlikely that logs across rivers constitute a more important source of food than logs fallen in the forest.

### Time to event

Another interesting analysis is the delay between detections of different individuals. Considering the scent-marking behavior, it is possible that individuals may be deterred from approaching or crossing the log if a conspecific, or a potential predator, has recently crossed. We follow the approach in [@Niedballa2019], who estimate the delay between detections at every site. For two species A and B, if there is avoidance or attraction of one with respect to the other, then the lag of A after B may be different than the lag of B after A. The effect is estimated using a linear model, with the log-transformed lags as a response variable.

#### Model

```{r}
#' First see if there are enough records, the paper by Niedballa
#' suggests you need at least 25 records for each species to get 
#' a proper estimation. I'll set a lower threshold of 20
d <- dat %>% group_by(station,species) %>% 
  filter(n()>20) %>% # filter species-stations with few records
  split(~station) 
spcomb <- lapply(d, function(x) outer(x$species,x$species,paste, sep="-"))
tdif <- lapply(d, function(x) outer(x$DateTime, x$DateTime, difftime, units = "mins"))
ddif <- data.frame(station = rep(names(spcomb), times = sapply(spcomb, length)), 
                sppair = unlist(spcomb), tdif = unlist(tdif)) %>% 
  filter(tdif > 0, # only positive differences
         tdif <= 24*60) %>% # only differences less than a day
  # mutate(spp = str_split(sppair, "-")) %>% 
  mutate(sp2 = gsub(replacement = "",x = sppair,pattern = "-{1}[[:alpha:]]+_?[[:alpha:]]+"),
         sp1 = gsub(replacement = "",x = sppair,pattern = "[[:alpha:]]+_?[[:alpha:]]+-{1}")) %>% 
  rowwise() %>%
  mutate(sporder1 = str_sort(c(sp2,sp1))[1],
         sporder2 = str_sort(c(sp2,sp1))[2],
         spp = paste(sporder1,sporder2, sep = "-")) %>% 
  filter(sp1!=sp2) %>% 
  split(~spp)
# Create model, glm with log link
lms <- lapply(ddif, \(x) glm(tdif~sppair, data = x, family = gaussian(link = "log")))
# # I tried to create also a glmm with station as a random variable, but there are stations with only one event so that doesn't work
# tte_glmms <- lapply(d, \(x) glmer(tdif~sppair+(1|station), data = x, family = gaussian(link = "log")))

# Select models with significant coefs
sig_mods <- which(sapply(lms, \(x) summary.glm(x)[["coefficients"]][2,4])<0.1)
# Get predicted times, degrees of freedom, p values and coefficient (link)
lapply(lms[sig_mods], \(x) cbind(spp = x$xlevels[[1]],
                                 t_pred = predict(x, newdata = x$xlevels, type = "response"),
                                 df = x$df.residual,
                                 p = summary(x)[["coefficients"]][2,4],
                                 beta = summary(x)[["coefficients"]][2,2]))
```

#### Observed time differences

The following is the mean observed time difference between detections of a focal species and any other species at a given site

```{r}
do.call(rbind, ddif) %>% group_by(sp2) %>%
  summarise(tmean=mean(tdif)/60, tsd = sd(tdif)/60) %>% 
  arrange(desc(tmean))

```

This is the difference for the same species

```{r}
dsame <- data.frame(station = rep(names(spcomb), times = sapply(spcomb, length)), 
                sppair = unlist(spcomb), tdif = unlist(tdif)) %>% 
  filter(tdif > 0, # only positive differences
         tdif <= 24*60) %>% # only differences less than a day
  # mutate(spp = str_split(sppair, "-")) %>% 
  mutate(sp2 = gsub(replacement = "",x = sppair,pattern = "-{1}[[:alpha:]]+_?[[:alpha:]]+"),
         sp1 = gsub(replacement = "",x = sppair,pattern = "[[:alpha:]]+_?[[:alpha:]]+-{1}"))  %>% 
  filter(sp1==sp2)

# plot
ggplot(dsame, aes(gsub(pattern = "-{1}[[:graph:]]+",x = sppair,replacement = ""), tdif))+geom_boxplot()+
  theme_classic(base_size = 12)+
  labs(y = "Time difference (minutes)",
       x = "Species")
```

These are the values for waiting time for the same species

```{r}
dsame %>% group_by(sp2) %>% summarise(tmean = mean(tdif)/60,twd = sd(tdif)/60)
```

We see that the interspecific waiting times are all higher than the respective conspecific wait times.

Plot this

```{r}
tdifcomb <- data.frame(station = rep(names(spcomb), times = sapply(spcomb, length)), 
                sppair = unlist(spcomb), tdif = unlist(tdif)) %>% 
  filter(tdif > 0, # only positive differences
         tdif <= 24*60) %>% # only differences less than a day
  # mutate(spp = str_split(sppair, "-")) %>% 
  mutate(sp2 = gsub(replacement = "",x = sppair,pattern = "-{1}[[:alpha:]]+_?[[:alpha:]]+"),
         sp1 = gsub(replacement = "",x = sppair,pattern = "[[:alpha:]]+_?[[:alpha:]]+-{1}"))
tdifcomb  %>% ggplot(aes(sp1,tdif))+
  geom_violin(aes(fill = (sp2==sp1)))+
  theme_classic(base_size = 16)+
  theme(axis.text.x = element_text(angle = -45,hjust=0))+
  labs(x = "Species 1", y = "Time difference (min)", fill = "Conspecific")
```

#### Model Summaries

```{r}
lapply(lms[sig_mods], summary.glm)
```

For most species pairs there was no evidence of temporal avoidance or attraction. We found evidence of shorter lags, suggestive of attraction, for skunks ($\beta=-0.69, t_{58}=-4.46, p < 0.01$) and opossums ($\beta=-0.34, t_{11}=2.03, p=0.067$) coming after coatis. Conversely, we found evidence suggestive of avoidance for common opossums coming after tamanduas ($\beta=-0.45, t_{17}=-1.93, p=0.071$), and of attraction for opossums coming after tayras ($\beta=0.74, t_{12}=3.18, p=0.0080$) and skunks ($\beta=0.70, t_{25}=2.23,p=0.035$).

This figure shows the observed lags in minutes for the species pairs that we found a difference for.

```{r}
# plot lags
do.call(rbind,d[sig_mods]) %>% 
  ggplot(aes(spp,tdif))+geom_boxplot(aes(fill = sppair))+
  theme_classic(base_size = 12)+
  labs(y = "Time difference (minutes)",
       x = "Species pairs")+
  theme(axis.text.x = element_text(hjust = 0, angle = -45))
```

I also determined the lag for individuals of the same species.

```{r}
# I could also buid a mixed model with these data
m <- glm(tdif~sppair, data = d, family = gaussian(link = "log"))
summary(m)
```

I created a linear model with the species as a predictor, and the time lag as a response. I used only the same-species times, to find the mean lag, and only times less than 12 h away to avoid bimodal distributions. The model estimates a mean lag of 175 minutes, or almost 3 hours, and there was no clear difference for any species.
